---
title: "R Ignite 2026 - Session 6: Bayesian Modeling with rstanarm"
output: html_document
---

## Purpose
Use familiar R formula syntax to fit, diagnose, and interpret a Bayesian logistic regression. No Stan-language programming is required.

```{r setup, echo=FALSE}

library(rstanarm)
library(bayesplot)
library(posterior)
options(mc.cores = parallel::detectCores())
set.seed(2026)
```

## 1. Data
```{r}
url <- "https://raw.githubusercontent.com/statOmics/PSLSData/main/NHANES.csv"
NHANES <- read.csv(url)
dat <- NHANES[!duplicated(NHANES$ID), ]
dat <- subset(dat, Age >= 20 & !is.na(Diabetes) & !is.na(Age) & !is.na(BMI) & !is.na(PhysActive))
dat$Diabetes01 <- ifelse(dat$Diabetes == "Yes", 1, 0)
dat$Age10 <- dat$Age / 10
dat$BMI5 <- dat$BMI / 5
```

## 2. Frequentist reference model
```{r}
fit_glm <- glm(Diabetes01 ~ Age10 + BMI5 + PhysActive, data=dat, family=binomial)
summary(fit_glm)
```

## 3. Bayesian model with rstanarm
```{r}
fit_bayes <- stan_glm(
  Diabetes01 ~ Age10 + BMI5 + PhysActive,
  data=dat, family=binomial(link="logit"),
  prior=normal(0,1,autoscale=FALSE),
  prior_intercept=normal(-2,1.5,autoscale=FALSE),
  chains=4, iter=2000, seed=2026, refresh=0
)
print(fit_bayes, digits=2)
prior_summary(fit_bayes)
```

## 4. Diagnostics
```{r}
summary(fit_bayes$stanfit)
mcmc_trace(as.array(fit_bayes), pars=c("Age10","BMI5"))
```

## 5. Credible intervals
```{r}
posterior_interval(fit_bayes, prob=.95)
```

## 6. Odds ratios and posterior probabilities
```{r}
draws <- as.matrix(fit_bayes)
OR_Age10 <- exp(draws[,"Age10"])
c(median=median(OR_Age10), quantile(OR_Age10,c(.025,.975)), prob_gt_1=mean(OR_Age10>1))
```

## 7. Posterior predictive check
```{r}
pp_check(fit_bayes, plotfun="bars", nreps=20)
```

## 8. Predicted probabilities
```{r}
new_patients <- data.frame(
  Age10=c(3,5,7), BMI5=c(5,6,7),
  PhysActive=factor(c("Yes","Yes","No"))
)
p_prob <- posterior_linpred(fit_bayes, newdata=new_patients, transform=TRUE)
apply(p_prob,2,median)
apply(p_prob,2,quantile,c(.025,.975))
```

## 9. Prior sensitivity (Extension Track)
```{r}
fit_wider <- update(fit_bayes, prior=normal(0,2,autoscale=FALSE))
fit_tighter <- update(fit_bayes, prior=normal(0,.5,autoscale=FALSE))
posterior_interval(fit_wider)
posterior_interval(fit_tighter)
```
