---
title: "R Ignite 2026 - Session 1 Hands-On Lab"
author: "Participant Name"
date: "September 11, 2026"
output: html_document
---

# My First Health Data Analysis

## Learning goals

By the end of this lab, you will be able to create a small health dataset, inspect it, calculate descriptive statistics, create a visualization, and write a short interpretation.

## 1. Create the data

```{r}
patient_data <- data.frame(
  ID = 1:10,
  Age = c(45, 50, 35, 60, 42, 29, 55, 48, 39, 67),
  BMI = c(28, 32, 24, 35, 27, 23, 31, 29, 26, 34),
  SystolicBP = c(122, 138, 118, 150, 128, 116, 142, 130, 124, 148),
  Diabetes = c("No", "Yes", "No", "Yes", "No", "No", "Yes", "No", "No", "Yes")
)
patient_data
```

**Checkpoint:** Before running `dim()`, predict how many rows and columns the dataset contains.

## 2. Inspect the data

```{r}
dim(patient_data)
names(patient_data)
str(patient_data)
summary(patient_data)
```

1. What type of variable is `Age`?  
2. What type of variable is `Diabetes`?  
3. What is the largest BMI?  

## 3. Descriptive statistics

```{r}
mean(patient_data$Age)
mean(patient_data$BMI)
sd(patient_data$BMI)
mean(patient_data$SystolicBP)
table(patient_data$Diabetes)
```

**Interpretation:** Write 1-2 sentences describing the typical age/BMI and the number of participants with diabetes.

## 4. First visualization

```{r}
hist(patient_data$BMI,
     main = "BMI Distribution",
     xlab = "BMI",
     ylab = "Frequency")
```

What does the distribution suggest?

## 5. Optional ggplot2 version

```{r, eval=FALSE}
# Run install.packages("ggplot2") only if the package is not installed.
library(ggplot2)

ggplot(patient_data, aes(x = BMI)) +
  geom_histogram(binwidth = 0.5, boundary = 0) +
  labs(title = "BMI Distribution", x = "BMI", y = "Count")
```

## 6. You do it

Change the graph to display `Age` instead of `BMI`. Change the title and x-axis label as well.

## 7. Extension track

```{r}
diabetes_group <- subset(patient_data, Diabetes == "Yes")
mean(diabetes_group$BMI)

aggregate(BMI ~ Diabetes, data = patient_data, FUN = mean)

plot(patient_data$Age, patient_data$SystolicBP,
     xlab = "Age", ylab = "Systolic BP",
     main = "Age and Systolic Blood Pressure")
```

**Extension question:** What additional analysis would you want before concluding that age is associated with blood pressure?

## 8. Debugging mini-challenge

The line below contains an error. Run it, read the message, and fix it.

```{r, error=TRUE}
mean(patient_data$bmi)
```

## Exit ticket

- One R command I can now use:  
- One thing I still find confusing:  
- One health/research dataset I would like to analyze in R:  
- Confidence using R now (1 = very low, 5 = very high):  
