---
title: "R Ignite 2026 - Session 3: Frequentist to Bayesian Inference"
output: html_document
---

## Health question
In a sample of 120 adults, 30 meet the study definition of hypertension.

```{r}
n <- 120
x <- 30
p_hat <- x/n
p_hat
```

## 1. Frequentist confidence interval
```{r}
prop.test(x, n, correct = FALSE)
```

## 2. Frequentist hypothesis test
H0: p = 0.20 versus H1: p != 0.20.

```{r}
prop.test(x, n, p = 0.20, correct = FALSE)
```

## 3. Bayesian prior
Likelihood: X | p ~ Binomial(n,p). Prior: p ~ Beta(a,b).

```{r}
a <- 2
b <- 8
a/(a+b)
```

## 4. Posterior updating
```{r}
a_post <- a + x
b_post <- b + n - x
post_mean <- a_post/(a_post+b_post)
post_ci <- qbeta(c(0.025, 0.975), a_post, b_post)
post_mean
post_ci
```

## 5. Posterior probability above 20%
```{r}
1 - pbeta(0.20, a_post, b_post)
```

## 6. Plot prior and posterior
```{r}
p <- seq(0, 0.6, length.out=500)
plot(p, dbeta(p,a,b), type="l", lwd=2,
     xlab="Hypertension prevalence p", ylab="Density",
     main="Prior and Posterior")
lines(p, dbeta(p,a_post,b_post), lwd=2, lty=2)
legend("topright", c("Prior Beta(2,8)","Posterior"), lty=c(1,2), lwd=2)
```

## 7. Prior sensitivity
```{r}
priors <- list(uniform=c(1,1), weak=c(2,8), stronger=c(20,80))
results <- do.call(rbind, lapply(priors, function(ab) {
  ap <- ab[1] + x
  bp <- ab[2] + n - x
  c(mean=ap/(ap+bp),
    lower=qbeta(.025,ap,bp),
    upper=qbeta(.975,ap,bp),
    prob_gt_20=1-pbeta(.20,ap,bp))
}))
results
```

## 8. Bridge to Session 6
This Beta-Binomial posterior is available analytically. Realistic Bayesian regression and hierarchical health models often require numerical posterior computation. Session 6 introduces MCMC for that purpose.
