---
title: "R Ignite 2026 - Session 2: Interactive R Primer with NHANES"
output:
  html_document:
    toc: true
    toc_float: true
---

## 1. Import directly from a URL

```{r}
url <- "https://raw.githubusercontent.com/statOmics/PSLSData/main/NHANES.csv"
NHANES <- read.csv(url)
# Save as a CSV file
#write.csv(NHANES, file = "HANES_R_Ignite.csv", row.names = FALSE)
dim(NHANES)
head(NHANES)
names(NHANES)
```

The expected raw dimensions are 10,000 rows and 76 columns. If the room network is unavailable, use a previously downloaded copy with `read.csv("NHANES.csv")`.

## 2. Inspect and create an adult person-level teaching dataset

```{r}
str(NHANES[c("ID","Gender","Age","Education","BMI","BPSysAve","Diabetes","HealthGen","PhysActive")])
sum(duplicated(NHANES$ID))

dat <- NHANES[!duplicated(NHANES$ID), ]
dat <- subset(dat, Age >= 20)
dim(dat)
```

Note: keeping the first row per repeated ID is a workshop simplification. In substantive research, investigate the reason for repeated records before choosing a dduplication rule.

## 3. Missingness and descriptive statistics

```{r}
vars <- c("BMI","Gender","Diabetes","HealthGen","BPSysAve","Age","PhysActive")
sapply(dat[vars], function(x) sum(is.na(x)))

summary(dat$BMI)
table(dat$Gender)
prop.table(table(dat$Diabetes))
with(dat, tapply(BMI, Gender, mean, na.rm=TRUE))
```

```{r}
hist(dat$BMI, main="BMI Distribution", xlab="BMI (kg/m^2)")
boxplot(BMI ~ Gender, data=dat, xlab="Gender", ylab="BMI (kg/m^2)")
```

## 4. Welch two-sample t-test

Research question: Does mean BMI differ between gender groups?

- H0: the population mean BMI is equal in the two groups.
- H1: the population means differ.

```{r}
tt <- t.test(BMI ~ Gender, data=dat)
tt
with(dat, tapply(BMI, Gender, mean, na.rm=TRUE))
```

## 5. Chi-square test of association

Research question: Is diabetes status associated with gender?

```{r}
tab <- with(dat, table(Gender, Diabetes, useNA="no"))
tab
prop.table(tab, margin=1)
chi <- chisq.test(tab)
chi
chi$expected
```

## 6. One-way ANOVA

Research question: Does mean average systolic blood pressure differ across self-reported general-health categories?

```{r}
anova_dat <- subset(dat, !is.na(BPSysAve) & !is.na(HealthGen))
anova_dat$HealthGen <- factor(anova_dat$HealthGen)
boxplot(BPSysAve ~ HealthGen, data=anova_dat, las=2,
        ylab="Average systolic BP", xlab="General health")

fit_aov <- aov(BPSysAve ~ HealthGen, data=anova_dat)
summary(fit_aov)
TukeyHSD(fit_aov)
```

### ANOVA diagnostics

```{r}
par(mfrow=c(1,2))
plot(fit_aov, which=1)
plot(fit_aov, which=2)
par(mfrow=c(1,1))
```

## 7. Logistic regression

Research question: How are age, BMI, gender and physical activity associated with the odds of diabetes?

```{r}
logit_dat <- subset(dat,
  !is.na(Diabetes) & !is.na(Age) & !is.na(BMI) &
  !is.na(Gender) & !is.na(PhysActive))

logit_dat$Diabetes01 <- ifelse(logit_dat$Diabetes == "Yes", 1, 0)
logit_dat$Gender <- factor(logit_dat$Gender)
logit_dat$PhysActive <- factor(logit_dat$PhysActive)

levels(logit_dat$Gender)
levels(logit_dat$PhysActive)
table(logit_dat$Diabetes01)

fit_logit <- glm(Diabetes01 ~ Age + BMI + Gender + PhysActive,
                 data=logit_dat, family=binomial)
summary(fit_logit)
```

### Odds ratios and Wald 95% confidence intervals

```{r}
OR_table <- cbind(
  OR = exp(coef(fit_logit)),
  exp(confint.default(fit_logit))
)
colnames(OR_table) <- c("OR","2.5%","97.5%")
round(OR_table, 2)
```

## 8. Predicted probabilities

```{r}
new_people <- data.frame(
  Age=c(30,50,70),
  BMI=c(24,30,36),
  Gender=factor(c("female","female","female"), levels=levels(logit_dat$Gender)),
  PhysActive=factor(c("Yes","Yes","No"), levels=levels(logit_dat$PhysActive))
)
new_people$Pr_Diabetes <- predict(fit_logit, newdata=new_people, type="response")
new_people
```

These are model-based teaching estimates, not clinical risk scores.

## 9. Extension track

```{r}
logit_dat$Age10 <- logit_dat$Age/10
fit_logit_10 <- glm(Diabetes01 ~ Age10 + BMI + Gender + PhysActive,
                    data=logit_dat, family=binomial)
exp(coef(fit_logit_10))
```

## 10. Exit questions

1. Why do we inspect and clean before hypothesis testing?
2. When would you choose a t-test, chi-square test, or ANOVA?
3. What does an odds ratio greater than 1 mean?
4. Why should this workshop analysis not be described as causal or as a population-weighted NHANES analysis?
